Von Neumann-Morgenstern and Savage Theorems for Causal Decision Making

### 2020-09-15 Artificial_Intelligence_Journal/Draft/AIJ_Draft.tex
Application: Formalizing an optimal action learning method for causal environments

In Section \ref{on_causality} we define Causality and state the differences between causal and associative information; having shown this, we state two of the most important decision making results, the von Neumann-Morgenstern and Savage Theorems, which rely on purely associative information; then, our objective is to present two analogous theorems: J. Pearl's result, Theorem \ref{causal_ut} which relies on a known causal model, and our main contribution: Theorem \ref{causal_savage}, which is able to use a set of \textit{beliefs} about possible causal models (i.e., a distribution) instead of a known one.

@@

We will consider $\Omega$ to be a family of admissible causal models; in this way,  $\omega \in \Omega$ being the true state of nature fixes a causal model which controls the environment in which the players make their choices. In classical Bayesian games, once $\omega \in \Omega$ is realized as the true state, then each player receives a signal $t_i=\tau_i (\omega)$ and the posterior belief $p_i(\omega | \tau^{-1}_i (t_i) )$ given by $p_i(\omega) / p_i (\tau^{-1}_i (t_i))$ if $\omega \in \tau^{-1}_i (t_i)$. In the case for causal bayesian games, we must consider both the probability $p_i$ of $\omega$ being the true state as well as the probability $p^\omega_i$ of observing a certain consequence when doing some action $a_i$ if $\omega$ is the true model.

@@

\section{Application: Formalizing an optimal action learning method for causal environments}
\label{optimal_action}
%lo que estoy planteando complementa estos trabajos que se hicieron de manera empirica
In recent papers by \cite{lattimoreNIPS2016,sen2017identifying} and \cite{gonzalez2018playing}, a decision maker is given the task to learn an optimal action from a series of learning rounds in a causal environment when the true causal model which controls the environment is unknown for the decision maker. \cite{lattimoreNIPS2016} assume that the conditional distributions for each variable given its parents is known, while \cite{sen2017identifying} assume as known only a part of the causal graphical model and allows for interventions in the unknown part of the model. \cite{gonzalez2018playing}  assume that the \textit{graphical structure} of such model is known. In their method, the decision maker holds beliefs about the parameters of the model, which are used in order to build a causal model that is used \textit{as if} it were the true one in order to obtain the \textit{best action} in each round.

\cite{gonzalez2018playing}, assume that the graphical structure of the causal model is known and use a Dirichlet distribution in order to sample the Conditional Probability Tables required to fully specify a Causal Graphical Model; once such model is specified the authors use an expression derived from Pearl's do-calculus in order to find a best action; the best action is defined in terms of the maximum expected utility as required by both the von Neumann-Morgenstern and Savage Theorems. Their method fits within our Theorem \ref{causal_savage} because in this paper the distribution $P_C(g)$ is a \textit{distribution over causal models with the same structure}, and when a model is specified from $P_C$, it is then used as if it were the true one in order to find the best action for such causal model. In the above mentioned paper, the possible outcomes are only $1$ and $0$, so once a model $g$ is sampled, the product $u(c)P_g(c|do(a))$ simply becomes $P(C=1 | do(a))$ which is stated only as an intuitive solution in the mentioned paper.

>>>>

In Section \ref{on_causality} we define Causality and state the differences between causal and associative information; having shown this, we state two of the most important decision making results, the von Neumann-Morgenstern and Savage Theorems, which rely on purely associative information; then, our objective is to present two analogous theorems: J. Pearl's result, Theorem \ref{causal_ut} which relies on a known causal model, and our main contribution: Theorem \ref{causal_savage}, which is able to use a set of \textit{beliefs} about possible causal models (i.e., a distribution) instead of a known model.

@@

We will consider $\Omega$ to be a family of admissible causal models; in this way,  $\omega \in \Omega$ being the true state of Nature fixes a causal model which controls the environment in which the players make their choices. In classical Bayesian games, once $\omega \in \Omega$ is realized as the true state, then each player receives a signal $t_i=\tau_i (\omega)$ and the posterior belief $p_i(\omega | \tau^{-1}_i (t_i) )$ given by $p_i(\omega) / p_i (\tau^{-1}_i (t_i))$ if $\omega \in \tau^{-1}_i (t_i)$. In the case for causal bayesian games, we must consider both the probability $p_i$ of $\omega$ being the true state as well as the probability $p^\omega_i$ of observing a certain consequence when doing some action $a_i$ if $\omega$ is the true model.

@@

\section{Application: Formalizing an optimal action learning method for causal environments}
\label{optimal_action}
In recent papers by \cite{lattimoreNIPS2016,sen2017identifying} and \cite{gonzalez2018playing}, a decision maker is given the task to learn an optimal action from a series of learning rounds in a causal environment when the causal model which controls the environment is unknown to the decision maker. \cite{lattimoreNIPS2016} assume that the conditional distributions for each variable given its parents is known, while \cite{sen2017identifying} assume as known only a part of the causal graphical model and allows for interventions in the unknown part of the model. Meanwhile, \cite{gonzalez2018playing} assume that the \textit{graphical structure} of the true model is known.

In the method described in \cite{gonzalez2018playing}, the decision maker holds probabilistic beliefs about the parameters of the model, which are used in order to build a \textit{local} causal model that is used \textit{as if} it were the true one in order to obtain the \textit{best action} in each round. \cite{gonzalez2018playing}, assume that the graphical structure of the causal model is known and use a Dirichlet distribution in order to sample the Conditional Probability Tables required to fully specify a Causal Graphical Model; once such model is specified the authors use an expression derived from Pearl's do-calculus in order to find a best action; the best action is defined in terms of the maximum expected utility as required by both the von Neumann-Morgenstern and Savage Theorems. Their method fits within our Theorem \ref{causal_savage} because in this paper the distribution $P_C(g)$ is a \textit{distribution over causal models with the same structure}, and when a model is specified from $P_C$, it is then used as if it were the true one in order to find the best action for such causal model. In the above mentioned paper, the possible outcomes are only $1$ and $0$, so once a model $g$ is sampled, the product $u(c)P_g(c|do(a))$ simply becomes $P(C=1 | do(a))$ which is stated only as an intuitive solution in the mentioned paper. Our result Theorem \ref{causal_savage} formalizes their method, thus providing solid ground for learning algorithms which rely on causal information.
